Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2021-10-05 analysis/paper/paper.Rmd
Updated text

## NIR - near infrared range
No distinct peaks occurred in the 600 – 1 000 nm band, and thus the analysis focused on the 1 000 – 2 500 nm band to lessen the impact of noise. In the case of NIRS most features relate to the presence of molecular water in the material [@hunt77; @sciuto_etal19], these mainly occur at around 1 400 nm and 1 900 nm and can be seen in the average spectra (fig. \@ref(fig:nir-spectra)). Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter could possibly be related to CO^3^ [@clark_etal90]. While the dark group features a higher overall absorbance the peaks are less distinct. The same can be said for the translucent and white group, where the water peak at 1 900 nm is the most pronounced. The absorbance in the translucent group is slightly higher in the lower band regions than compared to the white group, but with decreased absorbance in the higher regions. While a fifth peak at 2 450 nm can be difficult to discern in most cases....

The PCA demonstrates a similar pattern to the one observed in the visual band (fig. \@ref(fig:nir-pca)). Again the first two components explain most of the variance in the model, ~ 99.9 %. While the darker materials remain closely grouped in the right side of the score plot, the *light* cluster seems to split into two separate groups. The upper left group houses the main bulk of the *brecciated quartz*, as well as the almost translucent and opaque material with a beige tint. The bottom left group instead features the majority of the white *quartz-coloured* material.

@@

## Raman
There are two distinctive features present in the Raman data (fig. \@ref(fig:raman-spectra)). The first can be identified as the *quartz* peak occurring at 460 cm-1 [@gillet_etal90], which is present for all samples in the dataset. The second feature shows up as two peaks occurring at around 1 287 cm-1 and 1 598 cm-1, and dominate the spectra in terms of intensity for the darker samples. These peak positions fall within what is referred to as the D and G band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99]. The D band occurs as a result of disordered carbon structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21]. Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being 1 330 - 1 350 cm-1. It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.

A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ca. 695 cm-1 and 795 cm-1. Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks are present in measurements collected on, among others, albite, apatite, diabase and shale. Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm-1 band which has been related to Al-O-Si vibrations for the Kaolin group. Considering the diagenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset.

The resulting PCA model (fig. \@ref(fig:raman-pca)) shows a clear trend along the first component (81.5 %) related to the above two features. The measurements with mainly quartz peaks gravitate towards the positive side of the 1st component, and those where the graphite peaks are dominant towards the opposite side. It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more translucent objects mainly feature quartz peaks. As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks. 
The 2nd component (6.29 %) mainly seems to differentiate spectra with a lot of noise, where the objects were difficult to sample. In contrast, along the 3rd component those samples that feature more pronounced peaks around 700 - 800 cm-1 groups up separately from the *quartz group*.

@@

basic_plot2 <-
  fviz_pca_ind(raman.pca, axes = c(1,3), label="none")

#prepare labels for PCs
raman.pc1var <- round(summary(raman.pca)$importance[2,1]*100, digits=1)
raman.pc2var <- round(summary(raman.pca)$importance[2,2]*100, digits=1)
raman.pc3var <- round(summary(raman.pca)$importance[2,3]*100, digits=1)
raman.pc1lab <- paste0("PC1 (",as.character(raman.pc1var),"%)")
raman.pc2lab <- paste0("PC2 (",as.character(raman.pc2var),"%)")
raman.pc3lab <- paste0("PC3 (",as.character(raman.pc3var),"%)")

@@

The three first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ 86.5%. The loadings of the three components can be seen in figure \@ref(fig:xrf-loadings). In the score plot of P1 and P2, two large clusters emerges (fig. \@ref(fig:xrf-score-plot-p1-p2)). The loadings for ^13^Al, ^18^Sr, ^19^K, ^20^Ca, ^40^Zr, ^81^Ti group together in the top right quadrant whilst ^12^Mg, ^14^Si, ^15^P, ^16^S, ^23^V, and ^56^Ba group in the bottom right. As seen in figure \@ref(fig:xrf-score-plot-p1-p2), ^26^Fe is the only element positioned on the left side of the model. Interestingly, the left cluster contains the samples with lower ^14^Si values and the right cluster the higher values. These clusters still remain if you were to exclude ^14^Si from the model, however, suggesting there are other elements of relevance to the clustering. The “low ^14^Si” cluster shows a trend along P2 of quartzite with a darker hue grouping towards the top. This may indicate a possible relation to the ^26^Fe content of the material, as a darker hue may be the result of inclusions of hematite, or possibly biotite. ^13^Al follows a similar trend with higher values recorded at the top of the two clusters. 
The *brecciated quartz* samples largely group together at the bottom of the “low ^14^Si” cluster, with a couple ending up towards the bottom of the “high ^14^Si” cluster. The white (N9) material is somewhat evenly distributed within the leftmost cluster. 
The group of measurements that are somewhat separated from the “low ^14^Si” cluster at the top are all but one from the rock painting site of *Finnforsberget*, which is located next to the Skellefte river. The *Finnforsberget* artefacts were sampled twice with object 491 having one measurement closer to the “low ^14^Si” group and the other the “high ^14^Si” group with ^13^Al estimated at 25 %. This artefact was sampled twice on each side, resulting in similar results both times.
One almost translucent sample also ends up among the darker material. Upon closer inspection this sample is extremely small, 3 by 1 mm, and also feature inclusions of dark particles. Both of these factors may have contributed to its placement in the model.

>>>>

## NIR - near infrared range
No distinct peaks occurred in the 600 – 1 000 nm band, and thus the analysis focused on the 1 000 – 2 500 nm band to lessen the impact of noise. In the case of NIRS similar features can be observed for all four colour groups, with darker materials exhibiting higher absorbance and less defined peaks. The presence of molecular water can be seen in the peaks at 1 400 nm and 1 900 nm  [@hunt77; @sciuto_etal19] (fig. \@ref(fig:nir-spectra)). Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter falls within one of the stronger CO^3^ bands [@clark_etal90]. Inspection of individual spectra for the *light* group, however, also reveals a 5th peak for some samples at 2 450 nm. Comparisons to spectra collected on muscovite ([@clark_etal90] p.12 675) shows similar peaks located at 2 200 nm, 2 350 nm, and 2 450 nm. It is therefore possibly that the 2 350 nm band in some cases corresponds to CO^3^ vibrations and in other cases Al-OH.

The PCA shows a strong trend along PC 1 that seems to correlate with the colour of the material(fig. \@ref(fig:nir-pca)). Darker quartzite cluster to the right with white and translucent samples grouping to the opposite side. While the white samples seem to gravitate to the bottom left of the model the samples that are somewhat translucent, such as the *brecciated quartz*, seem to cluster towards the top-left. A small set of 3 black samples at the top of the model separate themselves clearly from the rest of the data points. These three samples come from different sites and exhibit no distinguishing peaks in their spectra, making them an interesting set of outliers.

@@

## Raman
There are two distinctive features present in the Raman data (fig. \@ref(fig:raman-spectra)). The first can be identified as the *quartz* peak occurring at 460 cm-1 [@gillet_etal90], which is present for all samples in the dataset. The second feature shows up as two peaks occurring at around 1 287 cm-1 and 1 598 cm-1, and dominate the spectra in terms of intensity for the darker samples. These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99]. The D band occurs as a result of disordered C structures whilst the G band is indicative of sp^2^ C networks and caused by stretching in the C bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21]. Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being 1 330 - 1 350 cm-1. It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.

A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ca. 695 cm-1 and 795 cm-1. Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks are present in measurements collected on, among others, albite, apatite, diabase and shale. Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm-1 band which has been related to Al-O-Si vibrations for the Kaolin group. Considering the petrogenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset.

The resulting PCA model (fig. \@ref(fig:raman-pca)) shows a clear trend along PC 1 (81.5 %) related to the quartz and D/G peaks. PC 2 and PC 3, however, mainly describes the remaining noise left in the spectra after preprocessing. Interestingly, PC 4 does catch the Al-O-Si peaks around the  700 cm-1 - 800 cm-1 band, which helps in separating out those samples where this feature is present. The measurements with mainly quartz peaks gravitates towards the positive side of the PC 1, and those where the graphite peaks are dominant towards the opposite side. It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more translucent objects mainly feature quartz peaks. As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks. When PC 4 is added the samples with quartz and Al-O-Si peaks gravitate towards the bottom left, separating themselves from the *quartz* group.

@@

basic_plot2 <-
  fviz_pca_ind(raman.pca, axes = c(1,4), label="none")

#prepare labels for PCs
raman.pc1var <- round(summary(raman.pca)$importance[2,1]*100, digits=1)
raman.pc2var <- round(summary(raman.pca)$importance[2,2]*100, digits=1)
raman.pc4var <- round(summary(raman.pca)$importance[2,4]*100, digits=1)
raman.pc1lab <- paste0("PC1 (",as.character(raman.pc1var),"%)")
raman.pc2lab <- paste0("PC2 (",as.character(raman.pc2var),"%)")
raman.pc4lab <- paste0("PC4 (",as.character(raman.pc4var),"%)")

@@

The three first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ 84.7%. The loadings of the three components can be seen in figure \@ref(fig:xrf-loadings). In the score plot of PC 1 and PC 2, two large clusters can be seen (fig. \@ref(fig:xrf-score-plot-p1-p2)). Interestingly, the smaller cluster on the left side in the model includes the samples with a SiO^2^ content higher than 50%, whilst the larger cluster includes those with lower SiO^2^ content. It is not ^14^Si that is responsible for this clustering, however, but rather ^26^Fe and ^40^Zr that drives this. The “low ^14^Si” cluster shows a trend along PC 2 of quartzite with a darker hue grouping towards the bottom. This may indicate a possible relation to the ^26^Fe content of the material as ^26^Fe-bearing minerals such as hematite can contribute to the dark colour in quartzite ([@howard05] p.711). The *brecciated quartz* samples largely group together at the top of the “low ^14^Si” cluster, with a couple ending up towards closer to the “high ^14^Si” cluster. The white (N9) material is somewhat evenly distributed within the right cluster.

The group of measurements that are somewhat separated from the “low ^14^Si” cluster at the bottom are all but one from the rock painting site of *Finnforsberget*, which is located next to the Skellefte river. The *Finnforsberget* artefacts were sampled twice with object 491 having one measurement closer to the “low ^14^Si” group and the other the “high ^14^Si” group with ^13^Al estimated at 25 %. This artefact was sampled twice on each side, resulting in similar results both times.

One almost translucent sample also ends up in the "high ^14^Si" cluster among the darker material. Upon closer inspection this sample is small, ~ 3 by 1 mm, and also feature inclusions of dark particles. Both of these factors may have contributed to its placement in the model.
